---
title: "Red Wine EDA"
author: "Mark Trapani"
date: "18 October 2018"
output:
  html_document: default
  pdf_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, error = TRUE)
mydata <- read.csv('wineQualityReds.csv')
```

This report explores a dataset which features 11 variables on the chemical 
properties of red wine with each entry having a grade rating between 
0 (very bad) to 10 (very excellent).

## Dataframe structure

```{r echo=FALSE}
col <- ncol(mydata) 
row <- nrow(mydata)

print(sprintf("Dataset has %s columns and %s rows", col,row))

str(mydata)
summary(mydata)

```

## Univariate Plots
 
To better understand the variables we're working with I am going to plot the 
distribution of each variable. The shape of each distribution should give an 
indication of what to expect when plotting different variables together. 

```{r echo=FALSE, Univariate_Plots1}
library(ggplot2)
ggplot(data = mydata, aes(x = quality)) +
  geom_bar(width = 0.5)

#changing wine quality to a factor rating
mydata$quality <- factor(mydata$quality, ordered = T)

mydata$rating <- ifelse(mydata$quality < 5, 'bad', 
                        ifelse(mydata$quality < 7, 'average', 'good'))

mydata$rating <- ordered(mydata$rating, levels = c('bad', 'average', 'good'))

ggplot(data = mydata, aes(x = rating)) +
  geom_bar(width = 0.5)

```

Looking at both graphs we can see how the majority of the wines we have in the
dataframe are average quality wines. This raises concerns on the accuracy of 
the data at hand and will make it diffuclt to build a model for the wine quality
due to the low count for bad and good ratings which act like outliers.


```{r echo=FALSE, Univariate_Plots2}

#creating new Total Acidity variable
mydata$total.acidity <- mydata$fixed.acidity + mydata$volatile.acidity

library(gridExtra)
p1 <- qplot(mydata$fixed.acidity, binwidth = 0.1, xlab = "Fixed Acidity")
p2 <- qplot(mydata$volatile.acidity, binwidth = 0.02, xlab = "Volatile Acidity")
p3 <- qplot(mydata$total.acidity, binwidth = 0.05, xlab = "Total Acidity")
p4 <- qplot(mydata$citric.acid, binwidth = 0.02, xlab = "Citric Acid")
p5 <- qplot(mydata$residual.sugar, binwidth = 1, xlab = "Residual Sugar")
p6 <- qplot(mydata$chlorides, binwidth = 0.02, xlab = "Chlorides")
p7 <- qplot(mydata$free.sulfur.dioxide, binwidth = 1, xlab = "Free Sulfur Dioxide")
p8 <- qplot(mydata$total.sulfur.dioxide, binwidth = 1.2, xlab = "Total Sulfur Dioxide")
p9 <- qplot(mydata$density, binwidth = 0.0005, xlab = "Density")
p10 <- qplot(mydata$pH, binwidth = 0.05, xlab = "pH")
p11 <- qplot(mydata$sulphates, binwidth = 0.05, xlab = "Sulphates")
p12 <- qplot(mydata$alcohol, binwidth = 0.5, xlab = "Alcohol")

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol = 4)

```

At first glance we can observe how most of the graphs have outliers so we need
to look at each variable separately to fully understand the data.


```{r echo=FALSE, warning=F, Functions}

#Outliers function to return instances above Q3 + 1.5IQR or below Q1 - 1.5IQR
remove_outliers <- function(x, na.rm = TRUE, ...) {
qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
H <- 1.5 * IQR(x, na.rm = na.rm)
y <- x
y[x < (qnt[1] - H)] <- NA
y[x > (qnt[2] + H)] <- NA
y
}


```

```{r echo=FALSE, warning=F, Univariate_Plots3}


grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(fixed.acidity))) +
               ylab("Fixed Acidity") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(4,14)),
ggplot(data = mydata, aes(x = remove_outliers(fixed.acidity))) +
  geom_histogram(binwidth = 0.4) +
  xlab("Fixed Acidity") +
  scale_x_continuous(lim = c(4,13)),ncol = 2)


```

The Fixed Acidity chart shows a right skewed distrubtion with a median of 
approximately 7.7 when removing the outliers.

```{r echo=FALSE, warning=F, Univariate_Plots4}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(volatile.acidity) )) +
               ylab("Volatile Acidity") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0.1,1.1)),
ggplot(data = mydata, aes(x = remove_outliers(volatile.acidity) )) +
  geom_histogram(binwidth = 0.03) +
  xlab("Volatile Acidity") +
  scale_x_continuous(lim = c(0.1,1.1)),ncol = 2)

```


The Volatile Acidity distribution is bimodal with peaks at 0.4 and 0.6.


```{r echo=FALSE, warning=F, Univariate_Plots5}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(total.acidity) )) +
               ylab("Total Acidity") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(5,14)),
ggplot(data = mydata, aes(x = remove_outliers(total.acidity) )) +
  geom_histogram(binwidth = 0.5) +
  xlab("Total Acidity") +
  scale_x_continuous(lim = c(5,14)),ncol = 2)

```

The Total Acidity has a positively skewed distribution after eliminating high
and low outliers.


```{r echo=FALSE, warning=F, Univariate_Plots6}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(citric.acid) )) +
               ylab("Citric Acid") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0.1,0.8)),
ggplot(data = mydata, aes(x = remove_outliers(citric.acid)) ) +
  geom_histogram(binwidth = 0.01) +
  xlab("Citric Acid") +
  ylim(0,75) +
  scale_x_continuous(lim = c(0,0.8)),ncol = 2)

```


Citric Acid data has a very unfamiliar shape since we're missing more than 
400 rows which resulted in a relatively rectangular distribution.


```{r echo=FALSE, warning=F, Univariate_Plots7}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(residual.sugar) )) +
               ylab("Residual Sugar") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(1,4)),
ggplot(data = mydata, aes(x = remove_outliers(residual.sugar) )) +
  geom_histogram(binwidth = 0.12) +
  xlab("Residual Sugar") +
  scale_x_continuous(lim = c(0.5,4)),ncol = 2)

```

The Residual Sugar distribution is also right skewed with a high solitary peak 
at 2.


```{r echo=FALSE, warning=F, Univariate_Plots8}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(chlorides) )) +
               ylab("Chlorides") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0.04,0.12)),
ggplot(data = mydata, aes(x = remove_outliers(chlorides) )) +
  geom_histogram(binwidth = 0.005) +
  xlab("Chlorides") +
  scale_x_continuous(lim = c(0.025,0.15)),ncol = 2)
```

Almost perfect normal distribution for Chloride variable after removing outliers.


```{r echo=FALSE, warning=F, Univariate_Plots9}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(free.sulfur.dioxide) )) +
               ylab("Free Sulfur Dioxide") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0,45)),
ggplot(data = mydata, aes(x = remove_outliers(free.sulfur.dioxide) )) +
  geom_histogram(binwidth = 1) +
  xlab("Free Sulfur Dioxide") +
  scale_x_continuous(lim = c(0,45)),ncol = 2)
```


Another positive skewed distribution for the Free Sulfur Dioxide variable.


```{r echo=FALSE, warning=F, Univariate_Plots10}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(total.sulfur.dioxide) )) +
               ylab("Total Sulfur Dioxide") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0,125)),
ggplot(data = mydata, aes(x = remove_outliers(total.sulfur.dioxide) )) +
  geom_histogram(binwidth = 2) +
  xlab("Total Sulfur Dioxide") +
  scale_x_continuous(lim = c(0,125)),ncol = 2)
```


Identical pattern followed by the Total Sulfur Dioxide variable.


```{r echo=FALSE, warning=F, Univariate_Plots11}

grid.arrange(ggplot(mydata, aes( x = 1, y = remove_outliers(density) )) +
               ylab("Density") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0.992,1.002)),
ggplot(data = mydata, aes(x = remove_outliers(density) )) +
  geom_histogram(binwidth = 0.001) +
  xlab("Density") +
  scale_x_continuous(lim = c(0.992,1.002)),ncol = 2)
```


We can observe an almost perfect normal distribution for the Density variable.


```{r echo=FALSE, warning=F, Univariate_Plots12}

grid.arrange(ggplot(mydata, aes(x = 1, y = remove_outliers(pH) )) +
               ylab("pH") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(2.9,3.7)),
ggplot(data = mydata, aes(x = remove_outliers(pH) )) +
  geom_histogram(binwidth = 0.05) +
  xlab("pH") +
  scale_x_continuous(lim = c(2.9,3.7)),ncol = 2)
```

A normal distribution can also be observed for the pH variable.

```{r echo=FALSE, warning=F, Univariate_Plots13}

grid.arrange(ggplot(mydata, aes(x = 1, y = remove_outliers(sulphates) )) +
               ylab("Sulphates") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(0.4,0.9)),
ggplot(data = mydata, aes(x = remove_outliers(sulphates) )) +
  geom_histogram(binwidth = 0.03) +
  xlab("Sulphates") +
  scale_x_continuous(lim = c(0.3,1)),ncol = 2)
```

Sulphate is slightly skewed to the right with a mean of almost 0.6.

```{r echo=FALSE, warning=F, Univariate_Plots14}

grid.arrange(ggplot(mydata, aes(x = 1, y = remove_outliers(alcohol) )) +
               ylab("Alcohol") +
               geom_jitter(alpha = 0.2 ) +
               geom_boxplot(alpha = 0.1, color = 'red' ) +
               scale_y_continuous(lim = c(8.7,12.5)),
ggplot(data = mydata, aes(x = remove_outliers(alcohol) )) +
  geom_histogram(binwidth = 0.2) +
  xlab("Alcohol") +
  scale_x_continuous(lim = c(8,14)),ncol = 2)
```

Final univariate plot shows the alcohol variable which has a right skewed 
distribution with a max value of 9.3.


## Univariate Plots Analysis

### What is the structure of your dataset?

The original dataset was made up of 1599 rows and 13 columns before adding an 
additional column called 'rating' where I've segmented wines as bad, average or 
good. This was done to get a broader understanding of the wines within this 
study. I've also added a Total Acidity column which is the sum of the fixed and 
volatile acidity. The categorical variable within our dataset is 'quality', 
whereas the others are numerical. 


### What is/are the main feature(s) of interest in your dataset?

Main features surrounding our analysis are the chemical and physical properties
of the wine to determine which of these effect wine quality.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

From my experience, the quality of wine is usually dependent on the acidity and 
sugar levels. Therefore I will have a closer look at the residual sugar and
total aciditiy variables. I will also include the pH level since it's a measure
of ripeness in relation to acidity and also effects the colour and lifespan 
of wine.


### Did you create any new variables from existing variables in the dataset?

I created a new variable called Total Acidity (Fixed + Volatile Acidity). Acids
in wine determines the overall taste and usually wines with a low acidity level 
have a very flat taste.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Since most of the data involved wines with a mid-range quaity rating, it is 
harder to determine what makes a good/bad wine. More data on the top and lower 
ranked wines would have made our analysis easier to work with.

After eliminating outliers we could still observe some skewed distributions 
particularly with variables used to determine taste. I thought there would
be a common middle range for residual sugar, free/total sulfur dioxide and 
sulphates. These all had a substantial amount of outliers. 


## Bivariate Plots

```{r echo=FALSE, warning=F, Bivariate_Plots1}

#remove unwanted columns for our correlation matrix


mydata_corr <-  mydata[, c("fixed.acidity",
                           "volatile.acidity",
                           "citric.acid",
                           "residual.sugar",
                           "chlorides",
                           "free.sulfur.dioxide",
                           "total.sulfur.dioxide",
                           "density",
                           "pH",
                           "sulphates",
                           "alcohol",
                           "quality",
                           "total.acidity"
                           )]


#changing quality variable to numeric
mydata_corr$quality <- as.numeric(as.character(mydata_corr$quality))

#plotting correlation matrix
corr <- cor(mydata_corr)
corrplot::corrplot(corr, method = "number",tl.cex=0.8,number.cex=0.5)

#correlation function
simple_cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}


```


1.  Looking at the above correlation matrix we can identify strong correlation
    between density and fixed/total acidity.
  
    
2.  Quality which is our main variable is correlated with alcohol and volatile
    acidity. There is also a low positive correlation with citric acids and
    sulphates.
    
3.  As expected, pH has a strong negative correlation with fixed acidity,
    however, there is a positive correlation with Volatile acidity which might
    be a classic case of Simpson's Paradox. This will be tested further on.
    
    
Now we will analyse different variables vs quality to get a better understanding 
of any exsiting relationships.

```{r echo=FALSE, warning=F, Bivariate_Plots2}
ggplot(data = mydata, aes(x = quality, y = fixed.acidity)) +
  ylab("Fixed Acidity") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)


correlations <- c(
  simple_cor_test(mydata$fixed.acidity, mydata$quality))
names(correlations) <- c('Fixed Acidity Correlation')

correlations

```

As illustrated in the above boxplots, both mean and median remain relatively
consistant even when looking at lower and higher quality grades. This shows that
fixed acidity has no effect on the overall quality.

```{r echo=FALSE, warning=F, Bivariate_Plots3}
ggplot(data = mydata, aes(x = quality, y = volatile.acidity)) +
  ylab("Volatile Acidity") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$volatile.acidity, mydata$quality))
names(correlations) <- c('Volatile Acidity Correlation')

correlations

```


On the other hand, looking at the relationship with volatile acidity we can
clearly notice that as the volatile acidity increases the quality of wine 
decreases.


```{r echo=FALSE, warning=F, Bivariate_Plots4}
ggplot(data = mydata, aes(x = quality, y = total.acidity)) +
  ylab("Total Acidity") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$total.acidity, mydata$quality))
names(correlations) <- c('Total Acidity Correlation')

correlations

```

Since most of the Total acidity is made up of Fixed Acidity the above boxplot
reflect the same pattern we had in the Fixed Acidity data.


```{r echo=FALSE, warning=F, Bivariate_Plots5}
ggplot(data = mydata, aes(x = quality, y = citric.acid)) +
  ylab("Citric Acid") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$citric.acid, mydata$quality))
names(correlations) <- c('Citric Acid Correlation')

correlations

```

Quality seems to have a positive correlation with citric acid. The above chart
shows a strong correlation which might be effected by the lack of low/high
quality wines data.

```{r echo=FALSE, warning=F, Bivariate_Plots6}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(residual.sugar))) +
  ylab("Residual Sugar") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$residual.sugar, mydata$quality))
names(correlations) <- c('Residual Sugar Correlation')

correlations

```

As observed in the correlation matrix, residual sugar has zero effect on the
quality of wine. One can notice how the mean values are stable across every 
quality grading. 

```{r echo=FALSE, warning=F, Bivariate_Plots7}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(chlorides))) +
  ylab("Chlorides") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$chlorides, mydata$quality))
names(correlations) <- c('Chlorides Correlation')

correlations

```

Small correlation present between Chlorides and Quality. When the quality
increases we notice the level of Chlorides going down gradually.

```{r echo=FALSE, warning=F, Bivariate_Plots8}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(free.sulfur.dioxide))) +
  ylab("Free Sulfur Dioxide") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$free.sulfur.dioxide, mydata$quality))
names(correlations) <- c('Free Sulfur Correlation')

correlations

```

Relationship between Free Sulfur Dioxide and Quality is quite strange since 
both low and high quality wine have a low level of Free Sulfur Dioxide. This
might mean two things. If the data lacks stats for low/high quality wines this
will end up giving a false representation of the relationship between both
variables, or either Free Sulfur Dioxide simply has no effect on the overall 
quality of wine.

```{r echo=FALSE, warning=F, Bivariate_Plots9}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(total.sulfur.dioxide))) +
  ylab("Total Sulfur Dioxide") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$total.sulfur.dioxide, mydata$quality))
names(correlations) <- c('Total Sulfur Dioxide Correlation')

correlations

```

As expected, since Total Sulfur Dioxide is a subset of Free Sulfur Dioxide we
get more or less the same pattern.


```{r echo=FALSE, warning=F, Bivariate_Plots10}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(density))) +
  ylab("Density") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$density, mydata$quality))
names(correlations) <- c('Density Correlation')

correlations

```

Another negative correlation might exist between Density and Quality. This 
might be driven by the alcohol content as observed in the correlation matrix.


```{r echo=FALSE, warning=F, Bivariate_Plots11}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(pH))) +
  ylab("pH") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$pH, mydata$quality))
names(correlations) <- c('pH Correlation')

correlations

```

A similar situation exists between pH and Quality. It might be interesting to 
investigate how pH behaves with the Total Acidity since lower pH usually shows
high levels of acids.

```{r echo=FALSE, warning=F, Bivariate_Plots12a}
ggplot(data = mydata, aes(x = total.acidity, y = pH)) +
  geom_point(alpha = 0.2) +
  scale_x_log10(breaks=seq(5,15,1)) +
  xlab("Total Acidity (Log Scale)") +
  geom_smooth(method="lm")

correlations <- c(
  simple_cor_test(mydata$total.acidity, mydata$pH))
names(correlations) <- c('Total Acidity Correlation with pH')

correlations
```


```{r echo=FALSE, warning=F, Bivariate_Plots12b}
ggplot(data = mydata, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.2) +
  scale_x_log10(breaks=seq(5,15,1)) +
  xlab("Fixed Acidity (Log Scale)") +
  geom_smooth(method="lm")

correlations <- c(
  simple_cor_test(mydata$fixed.acidity, mydata$pH))
names(correlations) <- c('Fixed Acidity Correlation with pH')

correlations
```


```{r echo=FALSE, warning=F, Bivariate_Plots12c}
ggplot(data = mydata, aes(x = volatile.acidity, y = pH)) +
  geom_point(alpha = 0.2) +
  scale_x_log10(breaks=seq(0.1,1,0.12)) +
  xlab("Volatile Acidity (Log Scale)") +
  geom_smooth(method="lm")

correlations <- c(
  simple_cor_test(mydata$volatile.acidity, mydata$pH))
names(correlations) <- c('Volatile Acidity Correlation with pH')

correlations

```

In these 3 scatter charts we plotted the behaviour between pH and acidity. One
can observe a contrasting behaviour in the relationship between Volatile Acidity
and pH when compared to Fixed/Total Acidity. This reversal behaviour might be a 
sign of Simpson's paradox which is investigated below.

```{r echo=FALSE, warning=F, results='hide', fig.keep='all', Bivariate_Plots13}
library(Simpsons)

simpsons <- Simpsons(volatile.acidity, pH, data=mydata)

```

As expected Simpson's paradox caused this reversal behaviour after comparing 3
data segments. 

```{r echo=FALSE, warning=F, Bivariate_Plots14}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(sulphates))) +
  ylab("Sulphates") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$sulphates, mydata$quality))
names(correlations) <- c('Sulphates Correlation')

correlations

```


Positive correlation exists between quality and sulphates. As quality in
wine increases the sulphates amount also becomes larger.


```{r echo=FALSE, warning=F, Bivariate_Plots15}
ggplot(data = mydata, aes(x = quality, y = remove_outliers(alcohol))) +
  ylab("Alcohol") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5)

correlations <- c(
  simple_cor_test(mydata$alcohol, mydata$quality))
names(correlations) <- c('Alcohol Correlation')

correlations


LinearModel <- lm(as.numeric(quality) ~ alcohol,
                       data = mydata)
summary(LinearModel)


```

We have the same situation with the Alcohol variable which increases with higher
quality wines. Given the strong correlation between Alcohol and Quality it is
was worth looking at the stats from a linear model. The R-Squared value shows 
that alcohol only makes up 22% of the Quality variable.



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

It was interesting to observe how variables such as sugar and fixed acidity have
no effect on quality. When observing acidity data we noticed how volatile
acidity has a negative correlation with quality.

Better graded wines tend to have higher alcohol content together with a high
concentration of Citric Acid. Investigating more red wine data may give a 
clearer picture on why alcohol only effects the variance of wine quality by 22%.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

When looking at the chemical features of wine we noticed a strong negative
correlation between density and alcohol. Density also positively correlates with
Total Acidity. 

As expected, pH negatively correlates with Fixed Acidity and Citric Acid,
however, we observed a low positive correlation with Volatile Acid. As shown
previously this was caused by Simpson's Paradox.

### What was the strongest relationship you found?

The strongest relationship is between Total Acidity and Density/pH. 


# Multivariate Plots Section

```{r echo=FALSE, warning=F, Multivariate_Plots1}
ggplot(data = mydata,
       aes(y = total.acidity, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) + 
  ylab("Total Acidity")

```

It seems that higher quality wine have higher alcohol and total acidity content.


```{r echo=FALSE, warning=F, Multivariate_Plots2}
ggplot(data = mydata,
       aes(y = density, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) + 
  ylab("Density")  

```

There doesn't seem to be any relationship between density and alcohol when 
trying to determine wine quality.

```{r echo=FALSE, warning=F, Multivariate_Plots3}
ggplot(data = mydata,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(0,1.5) +
  ylab("Sulphates")  

```

On the contrary, it appears that high quality wines have a high level of alcohol
and sulphates.


```{r echo=FALSE, warning=F, Multivariate_Plots4}
ggplot(data = mydata,
       aes(y = citric.acid, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(0,0.8) +
  ylab("Citric Acid")  

```

When plotting Citric Acid and Alcohol together we observe an upward trend for
both variables as the quality of the wine increases.

```{r echo=FALSE, warning=F, Multivariate_Plots5}
ggplot(data = mydata,
       aes(y = residual.sugar, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(0,4) +
  ylab("Residual Sugar")  

```

Residual Sugar and Alcohol have no relation when increasing wine rating.


```{r echo=FALSE, warning=F, Multivariate_Plots6}
ggplot(data = mydata,
       aes(y = total.sulfur.dioxide, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(0,100) +
  ylab("Total Sulfur Dioxide")  

```

No correlation exists between Total Sulfur Dioxide and Alochol.


```{r echo=FALSE, warning=F, Multivariate_Plots7}
ggplot(data = mydata,
       aes(y = pH, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(3,3.8) +
  ylab("pH")  

```

Higher quality wines seem to have a lower pH level and higher alcohol content.  


```{r echo=FALSE, warning=F, Multivariate_Plots8}
ggplot(data = mydata,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='seq',guide=guide_legend(title='Quality')) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ylim(0,1) +
  ylab("Volatile Acidity")  

```

Similar situation here where higher quality wines have lower volaile acidity
and higher alcohol content.


### Linear Model


```{r echo=FALSE, warning=F}

library(memisc)
m1 <- lm(as.numeric(quality) ~ alcohol, data = mydata)
m2 <- lm(as.numeric(quality) ~ alcohol + total.acidity, data = mydata)
m3 <- lm(as.numeric(quality) ~ alcohol + total.acidity + sulphates, data = mydata)
m4 <- lm(as.numeric(quality) ~ alcohol + total.acidity + sulphates + citric.acid, data = mydata)
m5 <- lm(as.numeric(quality) ~ alcohol + total.acidity + sulphates + citric.acid + volatile.acidity, data = mydata)

mtable(m1,m2,m3,m4,m5)
```

I performed a linear model analysis using the variables which have the highest
correlation with quality. The R-squared went marginally up to 34% which still 
shows that other variables might be at play to determine wine quality.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Citric Acid and Total Acidity strengthened each other since both seem to have a
positive correlation with wine quality. Sulphate also seems to have a positive 
effect on the quality levels of wine.

### Were there any interesting or surprising interactions between features?

The most interesting feature from the Multivariate analysis was that pH and 
Alcohol had no correlation with wine quality. Earlier we've seen that pH has a
strong negative correlation with Fixed/Total Acidity. In this analysis wine with 
higher quality has a larger amount of Fixed/Total Acidity.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created a good number of linear models to further understand data and how 
certain variables react against each other. Alcohol was the strongest variable 
to determine wine quality however since the R-squared was only 22% it shows 
that there are more variables to consider. When trying more variables in our 
linear models they failed to return a strong correlation (34%) with wine 
quality. This might be the effect of limited data for higher quality wines.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = mydata, aes(x = quality, y = alcohol)) +
  ylab("Alcohol (%)") +
  geom_jitter( alpha = .2) +
  geom_boxplot(alpha = .5,color = 'red') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 7, 
               size = 5) +
  xlab("Quality") +
  ggtitle("Alcohol vs Wine Quality")
```

### Description One

First plot shows how the alcohol content increases as wine quality levels go up.
This charts also shows a normal distribution since there is a very strong median
value. That being said, we've seen earlier that alcohol only contributes to 
around 22% in the variance of our outcome variable.


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(mydata, aes(x = alcohol, y = remove_outliers(sulphates), color = factor(quality))) +
  geom_jitter() +
  scale_color_brewer(name = "Quality") +
  ggtitle("Sulphates and Alcohol on Wine Quality") +
  xlab("Alcohol (%)") +
  ylab("potassium sulphate (g/dm3)")

```

### Description Two

Second graph shows the behaviour of Sulphate and Alcohol on the Quality of
Wine. We can see how higher Quality Wines (darker colour) are located mostly on
the top right of the chart which shows that higher ranked wines have high
Alcohol and Sulphate content.


### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(mydata, aes(x = alcohol, y = remove_outliers(volatile.acidity), color = factor(quality))) +
  geom_jitter() +
  scale_color_brewer(name = "Quality") +
  ggtitle("Volatile Acidity and Alcohol on Wine Quality") +
  xlab("Alcohol (%)") +
  ylab("Volatile Acidity (g/L)")

```

### Description Three

Third chart shows a contrasting pattern where higher quality wines are located 
on the bottom end which means that as we've seen earlier Volatile Acidity has a
negative correlation with wine quality. The chart shows a high concentration of 
light coloured markers towards the upper left hand side of the chart where 
Volatile Acidity is medium to high and Alochol content is very low which 
justifies our findings.

------

# Reflection

Our dataset included alot of chemical variables however alcohol was the one 
which stood out when trying to build a model which predicts wine quality. Other 
variables which to some extent correlated with wine quality were: total acidity, 
citric acid, sulphates and volatile acidity. All these variables were included
in a linear model which contributed to an R-Squared value of 0.344.

In this study the biggest issue was to get a low margin error to build a model 
which predicts the quality of wine. This might have been resolved if we had more 
data on higher ranked wines.

Some questions still remain after analysing this data. The first one relates to 
the variations in pH when comparing different acids. Secondly, further data may
answer the spikes we've seen in the citric acid chart.








